Global patterns of forest autotrophic carbon fluxes

### 2020-08-04 manuscript/manuscript draft.Rmd
update Fig. 2 caption, plus work on intro and discussion

Carbon cycling in Earth's forests provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations [@millennium_ecosystem_assessment_ecosystems_2005], while strongly influencing atmospheric carbon dioxide (CO~2~) and climate [@bonan_forests_2008]. 
Forests' autotrophic carbon fluxes--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance [@zak_plant_1994; @niedzialkowska_species_2010] and possibly diversity [@waide_relationship_1999; @chu_direct_2018]. 
They are linked to cycling of energy, water, and nutrients, and, critically, influence all carbon (C) stocks and define forest interactions with Earth's changing climate. 
Each year, over 69 Gt of C cycle through Earth's forests [@badgley_terrestrial_2019]-- a flux more than seven times greater that of recent anthropogenic fossil fuel emissions [9.5 Gt C yr^-1^; @friedlingstein_global_2019]. 
As atmospheric CO~2~ continues to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate [@cavaleri_urgent_2015; @rogelj_mitigation_2018]. 
However, our ability to draw general macroscopic conclusions regarding global variation in C cycling with respect to climate has been limited in that analyses have typically considered only one or a few variables at a time, have not always sufficiently parse related variables, and have often mixed forests that vary in stand age, disturbance history, and/or management status, all of which affect C cycling [@litton_carbon_2007; @gillman_latitude_2015; @simova_enigma_2017].

Forest C fluxes decrease with latitude [e.g., @luyssaert_co_2007; @piao_forest_2010; @gillman_latitude_2015; @li_mapping_2019]; but it remains unclear whether and how the shape of this relationship varies among fluxes. 
For instance, gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007]--or even a decrease [@huston_global_2009, but see @gillman_latitude_2015]. 
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016].
However, accurately comparing relationships of multiple forest C fluxes across latitude requires a large database with standardization for methodology, stand history, and management, which to date has not been possible.

@@

```{r echo=FALSE, out.width='100%', fig.cap = "Latitudinal trends in forest autotropic carbon flux. Lines of best fit are plotted according to the best model selected during analysis. All regressions are significant $(p<0.05)$. Each panel shows major C fluxes together with component fluxes. Also plotted are predicted trends in the major C fluxes based on the sum of component fluxes. 95\\% confidence intervals are plotted for the major flux for comparison with predicted trends. In (d), which shows three belowground fluxes, the major flux, total belowground carbon flux, has insufficient data (n=9) to support a regression", fig.pos='H'}

@@

This finding supports the continued focus on MAT as a primary--*albiet* not mechanistic--correlate of C fluxes. The effects of $MAT$ are modified by moisture availability, with reduced C fluxes under relatively dry conditions (*i.e.*, low precipitation; high vapour pressure deficit) and sometimes under very high precipitation (Figs. 3-4). **BBL: I feel like this is pretty repetitive of previous paragraph.** The observed positive interaction between MAT and MAP for $ANPP_{stem}$ on the global scale (Fig. 3) is consistent with an analysis showing a similar interaction for $ANPP$ in tropical forests, also with a cross-over point at ~20$\circ$C [@taylor_temperature_2017]. However, we detect no such interaction for $ANPP$ or most other C fluxes, and we find a contrasting negative interaction for $NPP$ (Fig. 3), suggesting that more data are required to sort out potential differences in the interactive effects of MAT and MAP on C fluxes in the tropics.

@@

Our analysis clarifies how forest autotrophic carbon fluxes vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of forest C cycling (Figs. 2-3). This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. **DISCUSS/CITE [Wu et al. 2020]** [@wu_accelerated_2020] However, increasing C flux rates are by no means universal [@rutishauser_testing_2020] (**MORE REFS**), likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. **(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)**
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on C cycling sets a foundation for understanding patterns of change.

>>>>

Carbon (C) cycling in Earth's forests provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations [@millennium_ecosystem_assessment_ecosystems_2005], while strongly influencing atmospheric carbon dioxide (CO~2~) and climate [@bonan_forests_2008]. 
Forests' autotrophic carbon fluxes--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance [@zak_plant_1994; @niedzialkowska_species_2010] and possibly diversity [@waide_relationship_1999; @chu_direct_2018]. 
They are linked to cycling of energy, water, and nutrients, and, critically, influence all C stocks and define forest interactions with Earth's changing climate. 
Each year, over 69 Gt of C cycle through Earth's forests [@badgley_terrestrial_2019]-- a flux more than seven times greater that of recent anthropogenic fossil fuel emissions [9.5 Gt C yr^-1^; @friedlingstein_global_2019]. 
As atmospheric CO~2~ continues to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate [@cavaleri_urgent_2015; @rogelj_mitigation_2018]. 
However, our understanding of the climate dependence of forest C cycling on a global scale has been limited by analyses typically considering only one or a few variables at a time, insufficient parsing of related variables, and the mixing of data from forests that vary in stand age, disturbance history, and/or management status, all of which affect C cycling [@litton_carbon_2007; @gillman_latitude_2015; @simova_enigma_2017].

Forest C fluxes decrease with latitude [e.g., @luyssaert_co_2007; @piao_forest_2010; @gillman_latitude_2015; @li_mapping_2019], but it remains unclear whether and how the shape of this relationship varies among fluxes. 
For instance, gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007]--or even a decrease [@huston_global_2009, but see @gillman_latitude_2015]. 
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016].
However, accurately comparing relationships of multiple forest C fluxes across latitude requires a large database with standardization for methodology, stand history, and management, which to date has not been possible.

@@

```{r echo=FALSE, out.width='100%', fig.cap = "Latitudinal trends in forest autotropic carbon flux. Plotted are linear models, all of which were significant $(p<0.05)$ and within dAIC=2 of the best model (for three fluxes, logarithmic fits were marginally better; Table S2). Each panel shows major C fluxes together with component fluxes. Also plotted are predicted trends in the major C fluxes based on the sum of component fluxes. 95\\% confidence intervals are plotted for the major flux for comparison with predicted trends. In (d), which shows three belowground fluxes, the major flux, total belowground carbon flux, has insufficient data (n=9) to support a regression", fig.pos='H'}

@@

This finding supports the continued focus on MAT as a primary--*albiet* not mechanistic--correlate of C fluxes. The effects of $MAT$ are modified by moisture availability, with reduced C fluxes under relatively dry conditions (*i.e.*, low precipitation; high vapour pressure deficit) and sometimes under very high precipitation (Figs. 3-4). **BBL: I feel like this is pretty repetitive of previous paragraph.** The observed positive interaction between MAT and MAP for $ANPP_{stem}$ on the global scale (Fig. 3) is consistent with an analysis showing a similar interaction for $ANPP$ in tropical forests, also with a cross-over point at ~20$\circ$C [@taylor_temperature_2017]. *(@sullivan_sensitivity_2020 show a decline in $ANPP_{stem}$ in sites with high max temperatures and limited moisture.)* However, we detect no such interaction for $ANPP$ or most other C fluxes, and we find a contrasting negative interaction for $NPP$ (Fig. 3), suggesting that more data are required to sort out potential differences in the interactive effects of MAT and MAP on C fluxes in the tropics.

@@

Our analysis clarifies how forest autotrophic carbon fluxes vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. 
Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of forest C cycling (Figs. 2-3).
This is consistent with observations of continental- to global-scale increases over time in $GPP$ [@li_mapping_2019] and $ANPP_{stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. 
**DISCUSS/CITE [Wu et al. 2020]** [@wu_accelerated_2020] 
However, increasing C flux rates are by no means universal [e.g., @rutishauser_testing_2020; @hubau_asynchronous_2020], likely because other factors are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. 
Carbon cycling is also likely to **(discuss new Sullivan paper that finds higher max temperatures associated to lower ANPP_stem)** [@sullivan_sensitivity_2020]
Moreover, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapted and acclimatized will not necessarily parallel responses across geographic gradients in climate. 
Nevertheless, understanding of the fundamental climatic controls on C cycling sets a firmer foundation for understanding forest C cycle responses to accellerating climate change.
